A Mechanised Library for Generic (Co)Programming

### 2023-12-31 paper/paper.tex
A draft of the motivations and intro

\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
\usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
\addbibresource{references.bib}
\usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{mathpartir}
\usepackage{bussproofs}
\usepackage{xparse}
\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{lipsum}
\usepackage{xargs}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{todonotes}
\usepackage{url}
\usepackage{xspace}
\usepackage{rotating}
\usepackage{quiver}
\usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
\addbibresource{references.bib}
\usepackage{array}   % for \newcolumntype macro
\newcolumntype{C}{>{$}c<{$}} % math-mode version of "l" column type
\newcolumntype{L}{>{$}l<{$}} % math-mode version of "l" column type

@@

\newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
\newcommand{\cata}[1]{\lbans #1 \rbans}
\newcommand{\ana}[1]{\llens #1 \rlens}
\newcommand{\fold}[1]{\catamor{#1}}
\newcommand{\unfold}[1]{\anamor{#1}}

\newcommand{\operator}[1]{\textsf{#1}}
\newcommand{\Alg}{\text{-Alg}}
\newcommand{\Free}{\text{Free\xspace}}

@@

\title{Hylomorphisms in Coq}

@@

\maketitle
\newcommand{\comp}{\cdot}
\section{Introduction}
Say that we want to prove that the two functions below are well-defined

@@

This is an instance of a more general phenonema whose argumet uses category theory.
\section{Adjoint Folds}
In order to transform $\operator{even-odd}$ into two functions $\operator{even}$ and
$\operator{odd}$ we are going to use something called an ``adjunction''.
In particular, we have two functors. The diagonal functor $\Delta : \CatC \to \CatC \times \CatC$ defined as
$\Delta X = (X,X)$ and the product functor $\times : \CatC \times \CatC$. These two functors form an adjunction:
\[\begin{tikzcd}
	{\CatC \times \CatC} & \CatC
	\arrow[""{name=0, anchor=center, inner sep=0}, "\Delta"', shift right=2, from=1-2, to=1-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\times"', shift right=2, from=1-1, to=1-2]
	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
    \arrow["\Delta\circ D\circ \times "', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
\end{tikzcd}
\]
More formally, this means there is a one-to-one correspondence of maps, i.e. every map $(A,A) \to (B_{1}, B_{2})$ corresponds univocally to a map of type $A \to B_{1} \times B_{2}$. In the isomorphism below we use $\ceiling{\cdot}$ and $\floor{\cdot}$ to denote the maps witnessing the isomorphism called respectively the right and left transpose of the adjunction:
\[
  \floor{\cdot} : \CatC \times \CatC ((A, A) , (B_{1}, B_{2}) \iso \CatC (A , B_{1} \times B_{2}): \ceiling{\cdot}
\]
the right transponse $\ceiling{\cdot}$ of the adjunction can be
defined using the projections:
\[
  \ceiling{g} = (\pi_{1} \cdot g,\pi_{2}\cdot g)
\]
and the floor can be defined as using the pair into the product:
\[
  \floor{f}(a) = \pair{f(a)}{f(a)}
\]

@@

In the following section we instantiate the adjoint fold with some instances
taken from the literature.
\end{document}

>>>>

\usepackage[utf8]{inputenc} % Required for inputting international characters
\usepackage[T1]{fontenc} % Output font encoding for international characters
\usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
\addbibresource{references.bib}
\usepackage{amsmath, amssymb, amsfonts, amsthm, stmaryrd}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{graphicx}
\usepackage{enumitem}
\usepackage{mathpartir}
\usepackage{bussproofs}
\usepackage{xparse}
\usepackage[usenames, dvipsnames]{xcolor}
\usepackage{lipsum}
\usepackage{xargs}
\usepackage{hyperref}
\usepackage{tikz-cd}
\usepackage{todonotes}
\usepackage{url}
\usepackage{xspace}
\usepackage{rotating}
\usepackage{quiver}
\usepackage[backend=bibtex,style=alphabetic,natbib=true]{biblatex}
\usepackage{minted}
\usepackage{newunicodechar}

\addbibresource{references.bib}
\usepackage{array}   % for \newcolumntype macro
\newcolumntype{C}{>{$}c<{$}} % math-mode version of "l" column type
\newcolumntype{L}{>{$}l<{$}} % math-mode version of "l" column type

@@

\newcommand{\anamor}[1]{{\llens\, #1\, \rlens}}
\newcommand{\catamor}[1]{\lbans\, #1\, \rbans}
\newcommand{\cata}[1]{\lbans #1 \rbans}
\newcommand{\ana}[1]{\llens #1 \rlens}
\newcommand{\fold}[1]{\catamor{#1}}
\newcommand{\unfold}[1]{\anamor{#1}}
\newcommand{\comp}{\cdot}
\newcommand{\operator}[1]{\textsf{#1}}
\newcommand{\Alg}{\text{-Alg}}
\newcommand{\Free}{\text{Free\xspace}}

@@

\newcommand{\haskell}[1]{\mintinline{haskell}{#1}}

\title{A Mechanised Library for Generic (Co)Programming}

@@

\maketitle
\section{Introduction}
Recursive definitions cannot be proven well-defined automatically. Modern proof
assistants like Coq or Agda provide a sound, but incomplete algorithm which
automatically checks for termination or productivity. For recursion, this is
done by trying to (automatically) infer which argument in the recursive call
gets smaller with respects to the original input argument. For productivity, the
algorithm checks that the (co)recursive call appears directly under a
constructor to make sure that this function always produce at least one element
after each recursive step.

As mentioned, these algorithms are incomplete meaning that some functions,
though well-defined, cannot be recognized so by the proof assistant.

One example of this fact is the quicksort algorithm given by the
\haskell{partition} and \haskell{quicksort} function below (written in Haskell
code):
\begin{minted}{haskell}
partition :: [a] -> ([a], a, [a])
partition (x:xs) = ((filter (x <=) xs), x, (filter (x >) xs))

quicksort :: [a] -> [a]
quicksort [] = []
quicksort ls = let (xs, x, ys) = partition ls in
                        (quicksort xs) ++ [x] ++ (quicksort ys)
\end{minted}
While both \haskell{partition} and the \haskell{quicksort} program using it are
well-defined functions a proof assistant such as Agda or Coq cannot accept such
definitions.

The argument can be given on pen-and-paper by showing the \haskell{partition} is
a function that always returns two sublists which are smaller than the one we
started with which means the repeated application of \haskell{partition} will
eventually lead to termination.

The quicksort algorithm is a \emph{divide-and-conquer} algorithm. Such
algorithms are structured in three parts: first, the input is broken down in
smaller sub-problems by means of a \emph{well-founded coalgebra}.Secondly, these
sub-problems are computed recursively and, finally, the results of the recursive
call are put together by a an algebra.

Practically, every program can be turned into an hylomorphism by means of the
conjugate rule~\cite{HinzeWG15}. This can be paraphrased by saying
\begin{center}
  ``\emph{every recursion scheme can be formalised as an hylomorphism}''
\end{center}

This has been a useful for structuring functional programs in a way that it
guarantees termination (or productivity) even in programming languages with
general recursion~\cite{DanielssonHJG06}.

With this work we intend to begin a study of recursion scheme beyond the world
of functional programming to the one of dependently typed languages such as a
the ones employed in proof assistants like Agda or Coq.

One of the challenges here is to be able to keep the theory of recursions
schemes as general as possible while maintaining its practical usefulness. In
order to do this, we have to make some design choices which differ a bit from
the general theory.  For instance, formalising the conjugate rule would require
us to mechanise a non-trivial fragment of category theory inside the prover's
logic.  Secondly, the general conjugate rule is a statement of correspondence of
arrows rather than a proof that these programs actually exists. In functional
programming languages like for instance Haskell this is not a problem, since
every functor is strong and has a fixed-point.  On the other hand, in a type
theory we have to restrict to those so-called polynomial functors. In this work
we do not make use of these assumptions. By restricting ourselves to polynomial
functors given in terms of containers~\cite{AbbottAG05} we are able to prove the
recursion schemes within the proof assistant and at the same time to maintain a
certain level of data abstraction.

First we prove that the hylo equation has a unique solution
\[
  x = a \comp F(x) \comp c
\]
for every recursive coalgebra $c$ for a polynomial $F$. Then we prove each
recursion scheme by reducing it to a hylo. The following recursion schemes can
all be proven by reducing them to an hylo: folds, mutual recursive functions,
recursion schemes from comonads and their respective coinductive dual.

The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion schemes in Coq
  \item We formalise algorithms for divide and conquer, dynamic programming and mutual recursion
  \item We validated optimisations in Coq and extract the resulting code to OCaml.
\end{itemize}

\section{Background on Adjoint Folds}
Say that we want to prove that the two functions below are well-defined

@@

This is an instance of a more general phenonema whose argumet uses category
theory.

\section{Adjoint Folds}
In order to transform $\operator{even-odd}$ into two functions $\operator{even}$
and $\operator{odd}$ we are going to use something called an ``adjunction''.  In
particular, we have two functors. The diagonal functor
$\Delta : \CatC \to \CatC \times \CatC$ defined as $\Delta X = (X,X)$ and the
product functor $\times : \CatC \times \CatC$. These two functors form an
adjunction:
\[\begin{tikzcd}
	{\CatC \times \CatC} & \CatC
	\arrow[""{name=0, anchor=center, inner sep=0}, "\Delta"', shift right=2, from=1-2, to=1-1]
	\arrow[""{name=1, anchor=center, inner sep=0}, "\times"', shift right=2, from=1-1, to=1-2]
	\arrow["\dashv"{anchor=center, rotate=-90}, draw=none, from=0, to=1]
    \arrow["D"', loop, distance=2em, in=35, out=325, from=1-2, to=1-2]
    \arrow["\Delta\circ D\circ \times "', loop, distance=2em, in=215, out=145, from=1-1, to=1-1]
\end{tikzcd}
\]
More formally, this means there is a one-to-one correspondence of maps, i.e.
every map $(A,A) \to (B_{1}, B_{2})$ corresponds univocally to a map of type
$A \to B_{1} \times B_{2}$. In the isomorphism below we use $\ceiling{\cdot}$
and $\floor{\cdot}$ to denote the maps witnessing the isomorphism called
respectively the right and left transpose of the adjunction:
\[
  \floor{\cdot} : \CatC \times \CatC ((A, A) , (B_{1}, B_{2}) \iso \CatC (A , B_{1} \times B_{2}): \ceiling{\cdot}
\]
the right transponse $\ceiling{\cdot}$ of the adjunction can be
defined using the projections:
\[
  \ceiling{g} = (\pi_{1} \cdot g,\pi_{2}\cdot g)
\]
and the floor can be defined as using the pair into the product:
\[
  \floor{f}(a) = \pair{f(a)}{f(a)}
\]

@@

In the following section we instantiate the adjoint fold with some instances
taken from the literature.

\printbibliography[heading=bibintoc]

\end{document}

%%% Local Variables:
%%% TeX-command-extra-options: "-shell-escape"
%%% End:
